---
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.align = "center", fig.width = 6)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# package loading
library(ggplot2)
library(gridExtra)
library(mFilter)
library(seasonal)
library(zoo)
source(file = "../R/write.zoo.dat.R")
```

```{r}
# data loading
data_ea <- read.csv.zoo("../data/raw_data/data_ea.csv",
  sep = ";",
  FUN = as.yearqtr,
  regular = TRUE
)
```

```{r}
# time series format
eonia_to <- data_ea$eonia_to |> as.ts()
estr <- data_ea$estr |> as.ts()
ipi_nsa <- data_ea$ipi_nsa |> as.ts()
defl_nsa <- data_ea$defl_nsa |> as.ts()
time <- zooreg(1:nrow(data_ea), order.by = index(data_ea), frequency = 4) |> as.ts()
```

# Preparations

## Variable names and descriptions

| Variable name | Description                                  |
|---------------|----------------------------------------------|
| `eonia`       | Eonia                                        |
| `eonia_to`    | Eonia with €STR methodology + 8.5 bp         |
| `estr`        | Euro short-term rate, €STR                   |
| `ipi_nsa`     | Industrial Production Index (2015 = 100) NSA |
| `defl_nsa`    | GDP deflator (2015 = 100) NSA                |
| `time`        | Time trend                                   |

## Annual inflation

### GDP deflator seasonal adjustment

Adjusting the GDP deflator `defl` via X-13ARIMA-SEATS, we obtain the seasonally adjusted GDP deflator `defl_sa`.

```{r}
defl <- seas(defl_nsa) |> final()
data_ea$defl <- defl
```

```{r echo=FALSE, fig.height=4.5, warning=FALSE}
p1 <- autoplot.zoo(cbind(defl_nsa, defl), facets = NULL) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  scale_color_manual(
    labels = c(
      "Original (defl_nsa)",
      "Adjusted (defl)"
    ),
    values = c("black", "red")
  ) +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

p2 <- autoplot.zoo(defl_nsa - defl) +
  theme_bw() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  labs(title = "Cycle") +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

grid.arrange(p1, p2,
  layout_matrix = cbind(c(1, 2)),
  heights = c(2, 1)
)
rm(p1, p2)
```

\newpage

### Seasonally adjusted GDP deflator percent change

Taking the percentage change of `defl`, we obtain the annual inflation rate created from the GDP deflator, `inf`.

```{r echo=FALSE}
inf <- (diff(as.zoo(defl), lag = 4, differences = 1, na.pad = T) / as.zoo(defl)) * 100
data_ea$inf <- inf
```

```{r echo=FALSE, fig.height=3, warning=FALSE}
p1 <- autoplot.zoo(defl) +
  theme_bw() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(title = "defl") +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

p2 <- autoplot.zoo(inf) +
  theme_bw() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(title = "inf") +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

grid.arrange(p1, p2,
  layout_matrix = rbind(c(1, 2))
)
rm(p1, p2)
```

## Output gap

### Industrial Production Index seasonal adjustment

Adjusting the Industrial Production Index `ipi_nsa` via X-13ARIMA-SEATS, we obtain the seasonally adjusted Industrial Production Index `ipi`.

```{r}
ipi <- seas(ipi_nsa) |> final()
data_ea$ipi <- ipi
```

```{r echo=FALSE, fig.height=4, warning=FALSE}
p1 <- autoplot.zoo(cbind(ipi_nsa, ipi), facets = NULL) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  scale_color_manual(
    labels = c(
      "Original (ipi_nsa)",
      "Adjusted (ipi)"
    ),
    values = c("black", "red")
  ) +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

p2 <- autoplot.zoo(ipi_nsa - ipi) +
  theme_bw() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  labs(title = "Cycle") +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

grid.arrange(p1, p2,
  layout_matrix = cbind(c(1, 2)),
  heights = c(2, 1)
)
rm(p1, p2)
```

### Seasonally adjusted Industrial Production Index Hodrick-Prescott filter

The creation of the output gap using the Industrial Production Index is done via obtaining the cyclical component of a Hodrick-Prescott filter with $\lambda = 1600$ on the seasonally adjusted Industrial Production Index `ipi`. The cyclical component obtained is multiplied by 100 on order to obtain the output gap `ogap`.

```{r}
ipi_hpf <- hpfilter(log(ipi), freq = 1600)
ogap <- ipi_hpf$cycle * 100
data_ea$ogap <- ogap
```

```{r echo=FALSE, fig.height=3.5, warning=FALSE}
p1 <- autoplot.zoo(cbind(log(ipi), ipi_hpf$trend), facets = NULL) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  scale_color_manual(
    labels = c(
      "log(ipi)",
      expression(paste("log(ipi), Hodrick-Prescott filter (", lambda, " = 1600)"))
    ),
    values = c(
      "black",
      "red"
    )
  ) +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

p2 <- autoplot.zoo(ipi_hpf$cycle) +
  theme_bw() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals") +
  scale_x_continuous(breaks = seq(1994, 2024, 2))

grid.arrange(p1, p2,
  layout_matrix = cbind(
    c(1, 2)
  ),
  heights = c(2, 1)
)
rm(p1, p2)
```

## Interest rate

In 2019, the European Central Bank (ECB), made the decision of deprecate the EONIA rate in favor of the euro short term rate (€STR).

In order to achieve an smooth transition from the EONIA to the €STR, the ECB released an EONIA computed with a new methodology that consist of the €STR plus a fixed spread of 8.5 basis points until January 3rd 2022. The EONIA with the original methodology was discontinued October 2nd 2019.

```{r echo=FALSE, fig.height=2.5, warning=FALSE}
autoplot.zoo(cbind(eonia_to, estr), facets = NULL) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  scale_color_manual(
    labels = c(
      "eonia_to",
      "estr"
    ),
    values = c(
      "black",
      "red"
    )
  ) +
  scale_x_continuous(breaks = seq(1994, 2024, 2))
```

```{r}
irate <- rbind(
  window(as.zoo(eonia_to), start = "1999 Q1", end = "2019 Q4"),
  window(as.zoo(estr), start = "2020 Q1")
) |> as.ts()
data_ea$irate <- irate
```

```{r echo=FALSE, fig.height=2}
autoplot.zoo(irate, facets = NULL) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1994, 2024, 2))
```

```{r}
# getting only the needed series and periods
data_ea_v1 <- merge(
  data_ea$irate,
  data_ea$inf,
  data_ea$ogap,
  data_ea$defl,
  data_ea$ipi,
  suffixes = c("irate", "inf", "ogap", "defl", "ipi")
) |>
  window(start = "1999 Q1", end = "2019 Q4")

data_ea_v1 <- round(data_ea_v1, 5)
```

```{r}
# write to .csv file
write.zoo(
  data_ea_v1,
  file = "../data/processed_data/data_ea_v1.csv",
  sep = ";"
)
```

```{r}
# write to .dat file
write.zoo.dat(data_ea_v1,
  file = "../data/processed_data/data_ea_v1.dat",
  comment = paste("Date", Sys.Date())
)
```
